---
title: "STAT 354 Project"
author: "Noah Schulhof"
date: today
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

# Load Packages

```{r}
library(tidyverse)
library(data.table)
```


# Data Preprocessing

```{r}
if (!(file.exists('data/counts.csv'))){
  dfs <- lapply(zip_files, function(z) {

    csv_names <- unzip(z, list = TRUE)$Name
    
    print(paste('Reading',
                csv_names[1]))
  
    lapply(csv_names, function(csv_name) {
  
      read_csv(
        unz(z, csv_name),
        col_select = started_at,
        show_col_types = FALSE
      )
  
    })

  })
  
  started_times <- rbindlist(unlist(dfs, recursive = FALSE))
  
  df <- as.data.frame(started_times)

  df$date <- as.Date(df$started_at)
  
  counts <- df %>% 
    group_by(date) %>% 
    summarize(count = n()) %>% 
    filter(date >= as.Date('2021-01-01')) %>% 
    arrange(date)
  
  write.csv(counts, 'data/counts.csv', row.names = F)
  
}
```

## Find number of daily rentals

```{r}
counts = read.csv('data/counts.csv') %>% 
  mutate(date = as.Date(date))

print(paste('Mean:',
      mean(counts$count)))

print(paste('Dispersion:',
            var(counts$count) / mean(counts$count)))
```

## Plot the resulting time series

```{r}
xt <- counts$count

png('plots/bike_counts.png', width = 800)

plot(counts$date, 
     xt,
     type = 'l',
     xlab = 'Date',
     ylab = 'Count',
     main = 'NYC Citi Bike Rental Count by Date')

dev.off()
```

## Examine the ACF of the raw and differenced time series

```{r}
png('plots/acf.png', width = 600)

acf(xt, main = 'ACF')

dev.off()
```

```{r}
xt_diff <- diff(xt)

png('plots/acf_diff.png', width = 600)

acf(xt_diff, main = 'ACF of Differenced Series')

dev.off()
```



# Fit SARIMA(p, d = 1, q) x (P, D = 0, Q) with S = 7

The differenced counts appear approximately stationary. We will fit an ARIMA(p, d = 1, q) model, and choose the values of p and q that minimize the AIC.

```{r, results='hide'}
train_window <- 730 # 2 years

init_train <- xt[1:train_window]

aic <- Inf

minAICp <- 0
minAICq <- 0
minAICP <- 0
minAICQ <- 0

vals <- seq(0, 1)

for (p in vals){
  for (q in vals){
    for (P in vals){
      for (Q in vals){
        
        fit <- tryCatch(
          arima(init_train,
                order = c(p, 1, q),
                seasonal = list(order = c(P, 0, Q),
                                period = 7),
                method = 'ML'),
          error = function(e) NULL
        )
        
        if (!is.null(fit)){
          
          new_aic <- fit$aic
        
          if (new_aic < aic){
            aic <- new_aic
            
            minAICp <- p
            minAICq <- q
            minAICP <- P
            minAICQ <- Q
          }
          
        }
    
        
      }
    }
      
  }
}
```


# Fit ACF(p, q)

```{r}
library(tscount)

LL <- tsglm(init_train,
            model = list(past_mean = NULL,
                         past_obs = NULL),
            distr = 'poisson')$logLik

maxLLp <- NULL
maxLLq <- NULL

vals <- c(NULL, seq(1, 5))

for (p in vals){
  for (q in vals){
      
    fit <- tsglm(init_train,
               model = list(past_mean = c(p, 7),
                            past_obs = c(q, 7)),
               distr = 'poisson',
               link = 'log',
               init.method = 'marginal',
               init.drop = F)
  
    if (fit$logLik > LL){
      LL <- fit$logLik
      
      maxLLp <- p
      maxLLq <- q
    }
      
  }
}
```


# Fit INGARCH with Negative Binomial

```{r}
LL <- tsglm(init_train,
            model = list(past_mean = NULL,
                         past_obs = NULL),
            distr = 'nbinom')$logLik

maxLLp2 <- NULL
maxLLq2 <- NULL

vals <- c(NULL, seq(1, 5))

for (p in vals){
  for (q in vals){
      
    fit <- tsglm(init_train,
               model = list(past_mean = c(p, 7),
                            past_obs = c(q, 7)),
               distr = 'nbinom',
               link = 'log',
               init.method = 'marginal',
               init.drop = F)
  
    if (fit$logLik > LL){
      LL <- fit$logLik
      
      maxLLp2 <- p
      maxLLq2 <- q
    }
      
  }
}
```


# Performance Comparison

```{r}
n <- length(xt)

test_start <- train_window + 1
```

## ARIMA predictions

```{r, results='hide'}
arima_preds <- rep(NA, n)

for(i in test_start:n){

  train <- xt[(i-train_window):(i-1)]

  fit <- arima(train,
                order = c(minAICp, 1, minAICq),
                seasonal = list(order = c(minAICP, 0, minAICQ),
                                period = 7),
                method = 'ML')
  
  pred <- predict(fit, n.ahead =1)

  arima_preds[i] <- pred$pred
}
```


## ACP predictions

```{r}
acp_preds <- rep(NA, n)

for (i in test_start:n){

  train <- xt[1:(i-1)]

  fit <- tsglm(
    train,
    model = list(
      past_mean = c(maxLLp,7),
      past_obs = c(maxLLq,7)
    ),
    distr = 'poisson',
    link = "log"
  )

  pred <- predict(fit, n.ahead = 1)

  acp_preds[i] <- pred$pred
}
```

## INGARCH Negative Binomial Predictions

```{r}
ingarch_preds <- rep(NA, n)

for (i in test_start:n){

  train <- xt[1:(i-1)]

  fit <- tsglm(
    train,
    model = list(
      past_mean = c(maxLLp2,7),
      past_obs = c(maxLLq2,7)
    ),
    distr = 'nbinom',
    link = "log"
  )

  pred <- predict(fit, n.ahead = 1)

  ingarch_preds[i] <- pred$pred
}
```


## Evaluate Results

### Fitted ACP order

```{r}
paste('p =',
      maxLLp,
      'q =',
      maxLLq)
```

### Fitted INGARCH order

```{r}
paste('p =',
      maxLLp2,
      'q =',
      maxLLq2)
```

### Fitted SARIMA order

```{r}
paste('p =',
      minAICp,
      'q =',
      minAICq,
      'P =',
      minAICP,
      'Q =',
      minAICQ)
```

### Combine Predictions

```{r}
results <- data.frame(
  date = counts$date,
  actual = xt,
  arima = arima_preds,
  acp = acp_preds,
  ingarch = ingarch_preds
)

results_test <- results[test_start:n, ]

head(results_test)
```

```{r}
evaluate_metrics <- function(actual, forecasted) {
  rmse <- sqrt(mean((actual - forecasted) ^ 2))
  mae <- mean(abs(actual - forecasted))
  mape <- mean(abs(actual - forecasted) / actual)
  
  data.frame(
    RMSE = round(rmse, 2),
    MAE  = round(mae, 2),
    MAPE = round(mape, 2)
  )
}
```



```{r}
results_df <- rbind(evaluate_metrics(results_test$actual, results_test$acp),
                    evaluate_metrics(results_test$actual, results_test$ingarch),
                    evaluate_metrics(results_test$actual, results_test$arima))

rownames(results_df) <- c('ACP', 'INGARCH (NegBin)', 'ARIMA')

results_df
```

```{r}
results_long <- results_test %>%
  pivot_longer(cols = c(actual, arima, acp, ingarch),
               names_to = "Model",
               values_to = "Count")

p <- results_long %>% 
  ggplot(aes(x = date, y = Count, color = Model)) +
  geom_line(linewidth = 0.25) +
  scale_color_manual(values = c(
    "actual" = "black",
    "arima" = "blue",
    "acp" = "red",
    "ingarch" = "green"
  ),
  labels = c("Actual", "ARIMA", "ACP", "INGARCH (Neg. Binom.)")) +
  labs(title = "Predicted vs. Actual",
       x = "Date",
       y = "Count",
       color = NULL) +
  theme_minimal(base_size = 14) +    
  theme(legend.position = "top")   

ggsave("plots/pred_actual.png", plot = p, width = 8, height = 5, dpi = 300)
```

```{r}
residuals <- results_test$actual - results_test$acp

png('plots/acp_res.png', width = 600)

acf(residuals,
    main = 'ACF of ACP Residuals')

dev.off()
```

```{r}
residuals <- results_test$actual - results_test$arima

png('plots/arima_res.png', width = 600)

acf(residuals,
    main = 'ACF of SARIMA Residuals')

dev.off()
```

```{r}
residuals <- results_test$actual - results_test$ingarch

png('plots/ingarch_res.png', width = 600)

acf(residuals,
    main = 'ACF of INGARCH (Neg. Binom.) Residuals')

dev.off()
```